### 2025-10-28 cluster-analysis.qmd
20251027 - additional detail

There are many approaches to cluster analysis, including model-based clustering, density-based clustering, centroid-based clustering, hierarchical clustering (aka connectivity-based clustering), etc.
In general, cluster analysis intends to maximize intracluster similarities and to minimize intercluster similarities [@Ramasubramanian2016].
Cluster analysis is a form of an [unsupervised approach to machine learning](#sec-machineLearningTypesUnsupervised) in which the groups are not known a priori.
There is a risk of reifying clusters identified in cluster analysis when, in some cases, the clusters may reflect artifacts of the analytic method rather than true distinct groups [@Everitt2011].

An overview of approaches to cluster analysis in `R` is provided by @Kassambara2017a.
In this chapter, we focus on examples using model-based clustering with the `R` package `mclust` [@R-mclust; @Scrucca2023_packages], which uses Gaussian finite mixture modeling.
The various types of `mclust` models are provided here:
<https://mclust-org.github.io/mclust/reference/mclustModelNames.html>.

>>>>

There are many approaches to cluster analysis, including model-based clustering, density-based clustering, centroid-based clustering, hierarchical clustering (aka connectivity-based clustering), etc.
In general, cluster analysis intends to maximize intracluster similarities and to minimize intercluster similarities [@Ramasubramanian2016].
Cluster analysis is a form of an [unsupervised approach to machine learning](#sec-machineLearningTypesUnsupervised) in which the groups are not known a priori.
There is a risk of reifying clusters identified in cluster analysis when, in some cases, the clusters may reflect artifacts of the analytic method rather than true distinct groups [@Everitt2011].
In general, the classification should be judged on the extent to which it is *useful* rather than on the extent to which it is true or false [@Everitt2011].

An overview of approaches to cluster analysis in `R` is provided by @Kassambara2017a.
In this chapter, we focus on examples using model-based clustering with the `R` package `mclust` [@R-mclust; @Scrucca2023_packages], which uses Gaussian finite mixture modeling.
Model-based clustering assumes the data are generated by an underlying statistical model, and the goal is to recover the structure of that model.
Model-based clustering represents the data as coming from a mixture of probability distributions.
The various types of `mclust` models are provided here:
<https://mclust-org.github.io/mclust/reference/mclustModelNames.html>.
